Covering Your Basis: Model-Agnostic Inference of LIGO-Virgos Binary Black Hole Popuation Properties

### 2022-09-19 src/tex/methods.tex
slowly filling out draft....

A common non-parametric method in statistical applications is the use of B-Splines or basis splines. A spline function of order $k$, 
is a piece-wise polynomial of order $k-1$ polynomials stitched together from defined ``knot'' locations across the domain. 
They provide a useful and cheap way to interpolate generically smooth functions from a finite sampling of ``knot'' heights. 
B-Splines of order $k$ are a set of order $k-1$ polynomials that form a complete basis for any spline function of order $k$. 
Given an array of knot locations, $\mathbf{t}$ or knot vector, there exists a single unique linear combination of B-Splines for 
every possible spline function with knots, $\mathbf{t}$. One can construct the B-Splines for a given knot vector using the Cox-de Boor 
recursion formula -- for knots $t_0$, $t_1$,...,$t_{i+k}$ the basis components are given by:
\begin{equation}
    B_{i,1}(x | \mathbf{t}) = 
    \begin{cases}
        1, & \text{if } t_i \leq x < t_{i+1} \\
        0, & \text{otherwise}
    \end{cases}
\end{equation}
\begin{equation}
    B_{i,k+1}(x | \mathbf{t}) = \omega_{i,k}(x)B_{i,k}(x) + \big[1-\omega_{i+1,k}(x)\big]B_{i+1,k}(x)
\end{equation}
\noindent Where $\omega_{i,k}$ above is defined as:
\begin{equation}
    \omega_{i,k}(x | \mathbf{t}) =
    \begin{cases}
        \frac{x-t_i}{t_{i+k}-t_i}, & t_{i+k} \neq t_i \\
        0, & \text{otherwise}
    \end{cases}
\end{equation}

A similar basis, the M-Spline basis, has desirable properties when looking to model a probability density function. Namely, 
$M_i \geq 0$ within $(t_i, t_i+k)$, zero elsewhere, and normalized to unity, $\int M_i(x)dx = 1$. The M-Spline basis only differs 
from B-Splines by a normalization factor, namely:
\begin{equation}\label{eq:MB_SplineRelation}
    M_{i,k} = \frac{k}{t_{i+k} - t_i} B_{i,k}
\end{equation}
\noindent We use this definition of the M-Spline basis for the models presented and applied in the later sections of this paper.

\subsection{Penalized Splines}
A common issue when using splines as a flexible fitting method is their sensitivity to the chosen number of knots, and their locations. 
Adding more knots knots will increase the natural a-priori variance in the spline function, while the space between knots can limit the 
resolution of features in the data the spline is capable of fitting. To ensure your spline based model is flexible enough to find sharp
features in data, one would want to add as many knots as densely as possible, but this comes with the sometimes un-wanted affect of the 
larger variance as mentioned above. To overcome this we resort to the frequentist penalized spline (PSpline) in which one applies a penalty
to the likelihood based on the difference of adjacent knot coeficients. Since the resulting linear combination of spline basis components is 
smoother, or flatter, when the basis coeficients are near each other, this has a natural smoothing affect to the spline function. Since we are 
using hierarchical Bayesian inferece as our statistical framework, we formulate this penalized likelihood with its Bayesian analog, by applying 
the stochastic differnece prior described in \bruce{CITE THIS PAPER}, which is equivalent to a Gaussian random walk prior where each succeeding 
coeficient is normally distributed about its neighboring coeficients. Thus our smoothing prior for our bayesian framework is:

\begin{equation}
    \tau_\lambda \sim \mathcal{G}(a, b)
\end{equation}
\begin{equation}
    p(\bm{c} | \tau_\lambda) \propto \exp{-\frac{1}{2} \tau_\lambda \bm{c}' \bm{D} \bm{c}}
\end{equation}

The smoothing prior removes the strong dependence on number and location of knots that arises with using splines. By adding in a 
very large number of knots such that your domain is densly populated with basis coeficients, the smoothing prior will smooth out regions where 
the data can be described smoothly, and relax when the data has features for the spline function to fit. In practice one would want to add knots 
of order the number of observations or more. One downside of this prior specification is a moderate dependence on choice of the b hyper-parameter for 
Gamma distribution prior on $\tau_\lambda$. We reduce this affect by following the method in \bruce{CITE BAYESIAN PSPLINES} where we construct 
a mixture model on a wide range of gamma distributions with different b hyper-parameters. Our prior is now defined as:

\begin{equation}
    \bm{p} \sim \mathcal{D}(n)
\end{equation}
\begin{equation}
    p(\tau_\lambda | \bm{p}) = \sum_i^m p_i \mathcal{G}(1, b_m)
\end{equation}
\begin{equation}
    p(\bm{c} | \tau_\lambda) \propto \exp{-\frac{1}{2} \tau_\lambda \bm{c}' \bm{D} \bm{c}}
\end{equation}

Additionally one can implement a spatially adaptive smoothing prior which allows for a smoothing prior
that can smoothly evolve over the domain of your spline function. This is done in practice by having the $\tau_\lambda$ parameter evolve 
across the knots. \bruce{CITE BAYESIAN PSPLINES}, formulates a spatially adaptive prior

\begin{equation}
    \bm{p} \sim \mathcal{D}(n)
\end{equation}
\begin{equation}
    p(\tau_\lambda | \bm{p}) = \sum_i^m p_i \mathcal{G}(1, b_m)
\end{equation}
\begin{equation}
    \lambda_k = \mathcal{G}(\omega, \omega)
\end{equation}
\begin{equation}
    p(\bm{c} | \tau_\lambda) \propto \exp{-\frac{1}{2} \tau_\lambda \bm{c}' \bm{D} \bm{\Lambda} \bm{D}' \bm{c}}
\end{equation}

Where above the matrix $\bm{\Lambda}$ is defined as: $\bm{\Lambda} = \mathrm{diag}\big(1,\lambda_0,...\lambda_k\big)$. This enforces a smooth evolution of the smoothing penalty 
across the domain that is being analyzed. This type of prior is especiially important when modeling distributions that span large orders of magnitude like 
the power-law like mass distributions that are commonly used.

>>>>

A common non-parametric method in statistical applications is the use of basis splines. A spline function of order $k$, 
is a piece-wise polynomial of order $k$ polynomials stitched together from defined ``knot'' locations across the domain. 
They provide a useful and cheap way to interpolate generically smooth functions from a finite sampling of ``knot'' heights. 
Basis splines of order $k$ are a set of order $k$ polynomials that form a complete basis for any spline function of order $k$. 
Given an array of knot locations, $\mathbf{t}$ or knot vector, there exists a single unique linear combination of B-Splines for 
every possible spline function with knots, $\mathbf{t}$. One can construct the B-Splines basis components for a given knot vector 
using the Cox-de Boor recursion formula. Given knots $t_0$, $t_1$,...,$t_{i+k}$ we start with the base case as:

\begin{equation}
    B_{i,1}(x | \mathbf{t}) = 
    \begin{cases}
        1, & \text{if } t_i \leq x < t_{i+1} \\
        0, & \text{otherwise}
    \end{cases}
\end{equation}

\noindent combined with the recursion relation:

\begin{equation}
    B_{i,k+1}(x | \mathbf{t}) = \omega_{i,k}(x | \mathbf{t})B_{i,k}(x | \mathbf{t}) + \\
                                \big[1-\omega_{i+1,k}(x | \mathbf{t})\big] B_{i+1,k}(x | \mathbf{t})
\end{equation}

\noindent where $\omega_{i,k}$ is defined as:

\begin{equation}
\omega_{i,k}(x | \mathbf{t}) =
\begin{cases}
    \frac{x-t_i}{t_{i+k}-t_i}, & t_{i+k} \neq t_i \\
    0, & \text{otherwise}
\end{cases}
\end{equation}

\noindent allowing one to build a basis of $i$ components that spans the vector space of order-$k$ spline functions interpolated 
from this knot vector. This is known as the ``B-Spline'' basis.

A similar basis, the M-Spline basis, has desirable properties when looking to model a probability density function. Namely, 
$M_i \geq 0$ within $(t_i, t_i+k)$, zero elsewhere, and normalized to unity, $\int M_i(x)dx = 1$. The M-Spline basis only differs 
from B-Splines by a normalization factor, namely:

\begin{equation}\label{eq:MB_SplineRelation}
M_{i,k} = \frac{k}{t_{i+k} - t_i} B_{i,k}
\end{equation}

\noindent We use this definition of the M-Spline basis for the models presented and applied in the later sections of this paper.

\subsection{Penalized Splines}
A common issue when using splines as a flexible fitting method is their sensitivity to the chosen number of knots, and their locations. 
Adding more knots knots will increase the natural a-priori variance in the spline function, while the space between knots can limit the 
resolution of features in the data the spline is capable of fitting. To ensure your spline based model is flexible enough to find sharp
features in data, one would want to add as many knots as densely as possible, but this comes with the sometimes un-wanted affect of the 
larger variance as mentioned above. To overcome this we resort to the frequentist penalized spline (PSpline) in which one applies a penalty
to the likelihood based on the difference of adjacent knot coeficients. Since the resulting linear combination of spline basis components is 
smoother, or flatter, when the basis coeficients are near each other, this has a natural smoothing affect to the spline function. Since we are 
using hierarchical Bayesian inferece as our statistical framework, we formulate this penalized likelihood with its Bayesian analog, by applying 
the stochastic differnece prior described in \bruce{CITE THIS PAPER}, which is equivalent to a Gaussian random walk prior where each succeeding 
coeficient is normally distributed about its neighboring coeficients. Thus the smoothing prior for $n$ degree's of freedom basis with a smoothing 
difference order of $r$ is defined as:

\begin{eqnarray}
\tau_\lambda \sim \mathcal{G}(a, b), \\
p(\bm{c} | \tau_\lambda) \propto \exp{-\frac{1}{2} \tau_\lambda \bm{c}^{\mathrm{T}} \bm{D_r}^{\mathrm{T}} \bm{D_r} \bm{c}}
\end{eqnarray}

\noindent Above $\bm{D_r}$ is the order-$r$ difference matrix, of shape $(n-r \times n)$ and $a$, $b$, shape parameters for the Gamma distribution, such that
the mean is at 1, with a large vairabce. This smoothing prior removes the strong dependence on number and location of knots that arises with using splines. 
By adding in a very large number of knots such that your domain is densly populated with basis coeficients, the smoothing prior will smooth out regions where 
the data can be described smoothly, and relax when the data has features for the spline function to fit. In practice one would want to add knots 
of order the number of observations or more. One downside of this prior specification is a moderate dependence on choice of the b hyper-parameter for 
Gamma distribution prior on $\tau_\lambda$. We reduce this affect by following the method in \bruce{CITE BAYESIAN PSPLINES} where we construct 
a mixture model on a wide range of gamma distributions with different b hyper-parameters. We can now replace the prior on $\tau_\lambda$ with:

\begin{eqnarray}
\bm{p} \sim \mathcal{D}(m), \\
p(\tau_\lambda | \bm{p}) = \sum_i^m p_i \mathcal{G}(1, b_m)
\end{eqnarray}

Additionally one can implement a spatially adaptive smoothing prior which allows for a smoothing prior
that can smoothly evolve over the domain of your spline function. This is done in practice by having the $\tau_\lambda$ parameter evolve smoothly
across the knots. \bruce{CITE BAYESIAN PSPLINES}, formulates a spatially adaptive prior in this way by adding in $n-r-1$ additional parameters that
effectively modulate the $\tau_\lambda$ across each difference. We now replace our prior on the coeficients with:

\begin{eqnarray}
\bm{\lambda} \sim \mathcal{G}(\omega, \omega), \\
\bm{\Lambda} \equiv \mathrm{diag}\big(1, \lambda_0,...,\lambda_{n-r-1}\big), \\
p(\bm{c} | \tau_\lambda) \propto \exp{-\frac{1}{2} \tau_\lambda \bm{c}^{\mathrm{T}} \bm{D_r}^{\mathrm{T}} \bm{\Lambda} \bm{D_r} \bm{c}}
\end{eqnarray}

Where above the matrix $\bm{\Lambda}$ a diagonal matrix of shape $(n-r \times n-r)$, and $\omega$ an arbitrarily small value such that the Gamma distribution 
prior on $\bm{\lambda}$ has a mean of 1 and an arbitrarily large variance. This enforces a smooth evolution of the smoothing penalty 
across the domain that is being analyzed. This type of prior is especiially important when modeling distributions that span large orders of magnitude like 
the power-law like mass distributions that are commonly used.

### 2022-09-19 src/tex/ms.tex
slowly filling out draft....

\title{Covering Your Basis: Model-Agnostic Inference of LIGO-Virgos Binary Black Hole Spin Distribution}

@@

\begin{abstract}
We introduce a non-parametric model built from basis splines, and apply it to infer the astrophysical mass distribution of 
binary black holes using GWTC-3.
\end{abstract}

>>>>

\title{Covering Your Basis: Model-Agnostic Inference of LIGO-Virgos Binary Black Hole Popuation Properties}

@@

\author{Zoheyr Doctor}
\affiliation{Center for Interdisciplinary Exploration and Research in Astrophysics (CIERA), Department of Physics and Astronomy, Northwestern
University, Evanston, IL 60201, USA}

\author{Jon Hornung}
\affiliation{Institute  for  Fundamental  Science, Department of Physics, University of Oregon, Eugene, OR 97403, USA}
\affiliation{Thurston High School, Springfield, OR 9747, USA}

\begin{abstract}
We introduce a non-parametric model built from basis splines, and apply it to infer the astrophysical distribution of 
binary black hole masses and spins using the 69 confident observations in GWTC-3. We find the same features previously recovered 
with similarly flexible models in the mass distribution, namely the peaks in merger rates at \~10\msun and \~35\msun. 
We apply our method to investigate whether the spin components are indendendenty distributed with a model agnositc way, where we find that 
the spin magnitude distribution seems to be quite consistently IID, with XX\% of black hole spins below XX at XX\% credibility. We report 
slight evidence of tilt distributions that do not peak at $\cos(\theta_i) = 1$ as previous models assumed, and slight evidence that 
the components are distributed differently, in contrast to the spin magnitude distribution. While we do not find signs of a significant
zero spinning sub population we do find that the data prefer not to force the support for exactly zero spinning black holes to 0 as 
the non-singular Beta distribution and \textsc{Default} model assumes. 
\end{abstract}
